Hydras \& Co.

### 2022-04-26 doc/JFLA2022/Slides/Slides.tex
work on JFLA slides

     \begin{block}{}
       \begin{itemize}
       \item In Coq, canonical sequences are defined by structural recursion on ordinal terms in Cantor normal form.
         \input{../../movies/snippets/Canon/canonExamplesb}
       \item Not the same definition as in K. \& S.'s article, but same abstract properties (monotony, limits, \dots).
       \end{itemize}
     \end{block}
\end{frame}

@@

    \begin{tikzpicture}[very thick, scale=0.25]
\node (alpha) at (0,0) {$\alpha$};
    \node (beta) at (40, 0){$\beta$};

@@

@@

\begin{frame}
  \begin{block}{}
    \begin{itemize}
    \item Let \textcolor{mathcolor}{$0<\alpha<\epsilon_0$}, and \textcolor{mathcolor}{$i\in\mathbb{N}$}.
      The least index \textcolor{mathcolor}{$l$} such that \textcolor{mathcolor}{$\alpha_l=0$} is  \textcolor{mathcolor}{$H'_\alpha(i)$} where \textcolor{mathcolor}{$H'_\alpha$} is defined by transfinite induction over \textcolor{mathcolor}{$\alpha$}.

      \begin{align}
        H'_0(i) & = i\\
        H'_\alpha(i) &= H'_{(\canonseq{\alpha}{i+1})}(i)  \quad\textit{if $\alpha$ is a limit ordinal}\\
        H'_{\alpha}(i) &=H'_\beta(i+1) \quad\textit{if $\alpha=\beta+1$}
      \end{align}

  \begin{block}{}
    The \textcolor{mathcolor}{$H'_\alpha$} family,  is a variant of the  \emph{Hardy hierarchy of rapidly growing functions}. For instance, \textcolor{mathcolor}{$H'_{\omega^3+3}(0)\geq2^{2^N}$}, where \textcolor{mathcolor}{$N=2^{70}-1$}.

    {\color{lookcolor} Such functions defy common programming intuition: no test is possible, only proofs!}
  \end{block}

@@

\begin{frame}
  \begin{block}{}
    The functions \textcolor{mathcolor}{$H'_\alpha\;(\alpha< \epsilon_0)$} are implemented with \textcolor{plugincolor}{coq-equations}.
    \vspace{5pt}
    \begin{footnotesize}
      \inputsnippets{../Hprime_HprimeDef}   
    \end{footnotesize}

  \begin{block}{}
    Using combinatorial lemmas by  K. \& S., we prove (by mutual  transfinite induction on \textcolor{mathcolor}{$\alpha$}):
    \begin{quote}
      \begin{itemize}
      \item \textcolor{mathcolor}{$H'_\alpha$} is strictly monotonous
      \item \textcolor{mathcolor}{$\forall i, i \leq H'_\alpha(i)$}
      \item   If \textcolor{mathcolor}{$\beta<\alpha$}, then \textcolor{mathcolor}{$H'_\beta(i)<H'_\alpha(i)$} for all but a finite number of $i$ (we say that \textcolor{mathcolor}{$H'_\alpha$}
        \emph{dominates} \textcolor{mathcolor}{$H_\beta$}).
      \end{itemize}
    \end{quote}

@@

\begin{frame}
  \begin{block}{Theorem}
    For any   \textcolor{mathcolor}{$\alpha\geq\omega^\omega$} the function \textcolor{mathcolor}{$H'_\alpha$} is not primitive recursive.
  \end{block}

  \begin{block}{}
    {\color{coqstylecolor}
      \begin{itemize}
      \item \emph{In order to prove this theorem, we chose to host and  maintain under \community Russel O'Connor's library \textcolor{plugincolor}{goedel}, which contains a formalization of primitive recursive functions.}

      \item    \emph{The interest of O'Connor's contribution led us to write a chapter on primitive recursive functions, with [counter-]examples and exercices.}

      \item \emph{Proving formally that a given function is not primitive recursive is a nice example of application of a complex (mutually recursive, dependently typed) induction principle. You have to take into account the set of PR functions of any arity.}

      \end{itemize}
    }
  \end{block}
\end{frame}

@@

    \begin{quote}
      For any \textcolor{mathcolor}{$n$}, and any primitive recursive function \textcolor{mathcolor}{$f$} of  arity \textcolor{mathcolor}{$n$}, there exists some natural number \textcolor{mathcolor}{$q$} such that the following inequality holds:
      {\color{mathcolor}
        \[
          \forall x_1,\dots,x_n, 
          f(x_1,\dots,\,x_n)\leq\textrm{Ack}(q,\textrm{max}(x_1,\dots,x_n))
        \]}
    \end{quote}
  \end{block}

    Instantiating with \textcolor{mathcolor}{$n=1$}, we obtain:

    \begin{quote}
      If \textcolor{mathcolor}{$H'_{\omega^\omega}$} were primitive recursive, there would exists some \textcolor{mathcolor}{$q$} such that, for any \textcolor{mathcolor}{$k$}, \textcolor{mathcolor}{$H'_{\omega^\omega}(k) \leq\textrm{Ack}(q,k)$}, which is false.
    \end{quote}

@@

\begin{frame}[fragile]
\begin{alltt}
{\color{lightblue}{(* begin snippet Ex42a:: no-out *)}}
Example Ex42: omega + 42 + omega^2 = omega^2. 
{\color{lightblue}{(* end snippet Ex42a *)}}
Proof.
  {\color{lightblue}{(* begin snippet Ex42b *)}}
  assert (HAP:= AP_phi0 2). {\color{lightblue}{(* .no-out *)}}
  elim  HAP; intros _ H0; apply H0; clear H0. 
  {\color{lightblue}{(* end snippet Ex42b *)}}
  {\color{lightblue}{(* begin snippet Ex42d *)}}
  Check AP_plus_closed. {\color{lightblue}{(* .unfold .no-goals *)}}
  {\color{lightblue}{(* end snippet Ex42d *)}}
   {\color{lightblue}{(* begin snippet Ex42c *)}}
  assert (Hlt: omega < omega^2) by 
      (rewrite omega_eqn; apply phi0_mono, finite_mono;
        auto with arith).
 {\color{lightblue}{ (* end  snippet Ex42c *)}}
\end{alltt}
\end{frame}

@@

\draw[->] (equations) -- (hydras) ;
\draw[->] (mathcomp) -- (gaia) ;
\draw[->] (mathcomp) -- (chains) ;
\draw[->] (paramcoq.south) -- (chains) ;
\draw[->] (pockcc) -- (goedelcc) ;
\end{tikzpicture}
}
\caption{\textcolor{lookcolor}{Genealogy} and dependencies for \Hydras packages.}
 %Dotted boxes represent historical Coq contributions, while regular boxes represent maintained \coq packages. Orange packages are maintained in the \Hydras GitHub repository, while blue packages are maintained in other \community repositories. Dotted lines represent \coq code ancestry, while regular lines represent direct code dependencies.}
  \label{fig:genealogy}
\end{figure}

>>>>

     \begin{block}{}
       \begin{itemize}
       \item In Coq, canonical sequences are defined by structural recursion on ordinal terms in Cantor normal form.

   {\color{mathcolor}
         $$\canonseq{\omega^{\omega^{\omega+1}+1}}{42}=
         \omega^{\omega^{\omega+1}}\times 42$$
         }

         \input{../../movies/snippets/Canon/canonExamplesb}
       \item Not the same definition as in K. \& S.'s article, but same abstract properties (monotony, limits, \dots).
       \end{itemize}
     \end{block}
\end{frame}

@@

          \label{fig:accessibility}
        \begin{tikzpicture}[very thick, scale=0.25]
\node (alpha) at (0,0) {$\alpha$};
    \node (beta) at (40, 0){$\beta$};

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{frame}
  \frametitle{Fast growing hierarchies of arithmetic functions}

  {\small
\textbf{Hardy Hierarchy (variant)}

 Let us define \textcolor{mathcolor}{$H'_\alpha(i)$} as the least $l$ such that \textcolor{mathcolor}{$\alpha_l=0$} (Figure page.~\pageref{fig:accessibility}). 
 {\color{mathcolor}
\begin{align}
  H'_0(i) & = i\\
  H'_{\beta+1}(i) &=H'_\beta(i+1) \\
  H'_\lambda(i) &= H'_{(\canonseq{\lambda}{i+1})}(i)  \quad\textit{($\lambda$ limit ordinal)} 
\end{align}}

\textbf{Wainer hierarchy (variant)}

We use also the following family in our proofs.

{\color{mathcolor}
\begin{align}
F_0(i)& =i+1\\
F_{\beta+1}(i)&= (F_\beta)^{(i+1)}(i)\\
F_\lambda(i) &= F_{(\canonseq{\lambda}{i})} (i) \quad \textit{($\lambda$ limit ordinal)}
\end{align}}}
\end{block}
\end{frame}

@@

\begin{frame}[fragile]

      \begin{block}{}
    The functions \textcolor{mathcolor}{$H'_\alpha, F_\alpha\;(\alpha< \epsilon_0)$} are implemented with \textcolor{plugincolor}{coq-equations}.
    \vspace{5pt}
    \begin{footnotesize}
      \inputsnippets{../Hprime_HprimeDef}   
    \end{footnotesize}

\begin{frame}[fragile]
  \begin{block}{Examples}

    Using the lemmas generated by \emph{coq-equations}, we prove minoration lemmas or exact formulas for given $\alpha$\,s:

    \begin{itemize}
    \item \input{../../movies/snippets/Hprime/HprimeOmegaCube3c}
      \input{../../movies/snippets/Hprime/HprimeOmegaCube3e}
     \item   \input{../../movies/snippets/Hprime/HprimeHexp2}
    \item
      \input{../../movies/snippets/Hprime/HprimePhi0OmegaClosed}

      \end{itemize}

@@

  \begin{block}{}
    Using combinatorial lemmas by  K. \& S., we prove (by mutual  transfinite induction on \textcolor{mathcolor}{$\alpha$}):
    \begin{quote}
      \begin{itemize}
      \item \textcolor{mathcolor}{$H'_\alpha$} is strictly monotonous
      \item If $\alpha>0$ then \textcolor{mathcolor}{$ i < H'_\alpha(i)$} for all $i$
        \item $H'_\alpha(i) < H'_{\alpha+1}(i) \; (i \not=0)$
      \item   If \textcolor{mathcolor}{$\alpha \rplus{n,n,\dots,n} \beta$} then \textcolor{mathcolor}{$H'_\beta(n)\leq H'_\alpha(n)$} 
      \end{itemize}
    \end{quote}

  \end{block}
  \begin{block}{Theorem}
    The function which computes the length of hydra battles
    (according to the initial index) is not primitive recursive
    (for $\alpha\geq \omega^\omega$).
    \vspace{4pt}

\emph{We use   Russel O'Connor's formalization of PR functions.}
  \end{block}

@@

% \begin{frame}

%   \begin{block}{}
%     {\color{coqstylecolor}
%       \begin{itemize}
%       \item \emph{In order to prove this theorem, we chose to host and  maintain under \community Russel O'Connor's library \textcolor{plugincolor}{goedel}, which contains a formalization of primitive recursive functions.}

%       \item    \emph{The interest of O'Connor's contribution led us to write a chapter on primitive recursive functions, with [counter-]examples and exercices.}

%       \item \emph{Proving formally that a given function is not primitive recursive is a nice example of application of a complex (mutually recursive, dependently typed) induction principle. You have to take into account the set of PR functions of any arity.}

%       \end{itemize}
%     }
%   \end{block}
% \end{frame}

@@

    \begin{quote}
      For any \textcolor{mathcolor}{$n$}, and any primitive recursive function \textcolor{mathcolor}{$f$} of  arity \textcolor{mathcolor}{$n$}, there exists some natural number \textcolor{mathcolor}{$q$} such that the following inequality holds:
      {\color{mathcolor}
        \[
          \forall x_1,\dots,x_n, 
          f(x_1,\dots,\,x_n)\leq\textrm{Ack}(q,\textrm{max}(x_1,\dots,x_n))
        \]}
    \end{quote}

       \input{../../movies/snippets/F_omega/lemmas}

@@

\begin{frame}[fragile]
\begin{alltt}
{\color{lightblue}{(* begin snippet Ex42a:: no-out *)}}
Example Ex42: omega + 42 + omega^2 = omega^2. 
{\color{lightblue}{(* end snippet Ex42a *)}}
Proof.
  {\color{lightblue}{(* begin snippet Ex42b *)}}
  assert (HAP:= AP_phi0 2). {\color{lightblue}{(* .no-out *)}}
  elim  HAP; intros _ H0; apply H0; clear H0. 
  {\color{lightblue}{(* end snippet Ex42b *)}}
  {\color{lightblue}{(* begin snippet Ex42d *)}}
  Check AP_plus_closed. {\color{lightblue}{(* .unfold .no-goals *)}}
  {\color{lightblue}{(* end snippet Ex42d *)}}
   {\color{lightblue}{(* begin snippet Ex42c *)}}
  assert (Hlt: omega < omega^2) by 
      (rewrite omega_eqn; apply phi0_mono, finite_mono;
        auto with arith).
 {\color{lightblue}{ (* end  snippet Ex42c *)}}
\end{alltt}

@@

\draw[->] (equations) -- (hydras) ;
\draw[->] (mathcomp) -- (gaia) ;
\draw[->] (mathcomp) -- (chains) ;
\draw[->] (paramcoq.south) -- (chains) ;
\draw[->] (pockcc) -- (goedelcc) ;
\end{tikzpicture}
}
\caption{\textcolor{lookcolor}{Genealogy} and dependencies for \Hydras packages.}
% Dotted boxes represent historical Coq contributions, while regular boxes represent maintained \coq packages. Orange packages are maintained in the \Hydras GitHub repository, while blue packages are maintained in other \community repositories. Dotted lines represent \coq code ancestry, while regular lines represent direct code dependencies.}

  \label{fig:genealogy}
\end{figure}

### 2022-04-26 doc/part-hydras.tex
changes to JFLA slides
work on JFLA slides

For instance \coq's computing facilities allow us to verify the equalities\linebreak 
\mathcolor{$\canonseq{\omega^\omega}{3} = \omega^3$} and
\mathcolor{$\canonseq{\omega^\omega*3}{42} = \omega^\omega*2 + \omega^{42}$}.

\input{movies/snippets/Canon/canonExamples}

@@

Note that this formula contains two occurrences of the functional \texttt{iterate}, the outer one is in fact a second-order iteration (on type \texttt{nat -> nat)}
and the inner one  first-order (on type \texttt{nat}).

@@

Thus,  our function \texttt{ord}  which enumerates the elements of $B$ is defined in a non-ambiguous way.
Let us quote the following theorems (see Library
\href{../theories/html/hydras.Schutte.Ordering_Functions.html}%
{\texttt{Schutte.Ordering\_Functions}} for more details).

>>>>

For instance \coq's computing facilities allow us to verify the equalities\linebreak 
\mathcolor{$\canonseq{\omega^\omega}{3} = \omega^3$} and
\mathcolor{$\canonseq{\omega^{\omega^{\omega+1}+1}}{42}=
 \omega^{\omega^{\omega+1}}\times 42$}

\input{movies/snippets/Canon/canonExamples}
\input{movies/snippets/Canon/canonExamplesb}

@@

Note that this formula contains two occurrences of the functional \texttt{iterate}, the second  one is in fact a second-order iteration (on type \texttt{nat -> nat)}
and the first one  first-order (on type \texttt{nat}).

@@

Thus,  our function \texttt{ord}  which enumerates the elements of $B$ is defined in a non-ambiguous way.
Let us quote the following theorems (see Library \linebreak
\href{../theories/html/hydras.Schutte.Ordering_Functions.html}%
{\texttt{Schutte.Ordering\_Functions}} for more details).
